%!TEX root = ../tobias_neumann_phd_thesis.tex

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
This thesis was the result of a lot of discussions and interactions with people with many different backgrounds and affiliations who I have to wholeheartedly thank for their contributions: \\
First, there is my supervisor Arndt who really facilitated a fast and uncomplicated integration into his group, giving me the freedom to pursue my work and boiling down bureaucracy to the necessary minimum. Without this, I would not have managed to pursue this thesis in anywhere close the same timeframe and efficiency. I would like to thank also all CIBIV members and the subgroup members for the continuous discussions and invaluable feedback on my progress. \\
Equally important to this thesis is my former mentor at the IMP Johannes Zuber who actively encouraged me to pursue this PhD and put a lot of energy and commitment into making it happen and in general taught me a lot about scientific rigor and out-of-the box thinking. Hannes was also the one who recruited me to work on the SLAMseq technology which brought me in contact with Stefan Ameres: He introduced me to cutting-edge transcriptomics and metabolic labelling which form the biological basis of this thesis. Stefan's lab was also exceptionally welcoming and spawned many fruitful collaborations up to this day. Stefan together with Hannes were detrimental for my career when they recruited me to Quantro Therapeutics to bring SLAMseq from basic research to drug discovery to ultimately help patients. \\
Many great collaborators and friends helped to get me out of the comfort zone and made my projects a success: Philipp as a great friend who I can always discuss algorithms with but also politics and life in general - and my constant Nanopore troubleshooting contact. Niko who relentlessly drove the splice\_sim development and is a great conceptional thinker who I sometimes wonder how on earth he handles so many project simultaneously. Veronika, who was instrumental to improve my manuscript writing and enabled me to handle a paper as corresponding author. Nina, who is a great technical expert and methodological thinker and amazing co-worker. I want to explicitly also thank Dieter and the board of Quantro Therapeutics for supporting this PhD even through a hectic startup phase. From an administrative point of view, I am forever indebted to Gerlinde who handled all my countless emails with patience and helped me through all the peculiarities of my PhD. \\
Thanks also to Lukas for teaching me a lot about cancer genetics and being integral to linking my bioinformatics effort with biological impact. A special thanks to Daniel who, starting as a Master's student, surpassed me in a ridiculously short amount of time in many areas and is now a crucial source of advise and a good friend. I would also never have managed any of this without Mihi who accompanied me along the way, starting from coaching me for the interview all the way to the end and being the best friend I could wish for. The discussions with her taught me most of what I know about wetlab work, about experimental thinking and showed me what a team of wetlab and computational scientists can achieve together. \\
Lastly, I would like to thank my family who gave me the room and time I needed to get this thesis over the finish line and dragged me out of Science when I got too much caught up with it.